---
title: "Project 1: Concentration risk tool"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
---

This section describes the concentration risk tool.. 

# Problem

Depending on the available information and the desired format, the cumulation risk given an address or a given location is showed on a map or in a table. Next to this a list with the coördinates with the highest cumulation risks within the Netherlands is given. 

The goal of the tool is to determine the locations with the highest cumulation risks (small-areas with a radius of 200 meters). Furthermore, the tool gives Delta Lloyd acceptants during the accepting process in a fast and simple way insight in the cumulation risk on addresses and locations in the Netherlands. 

## Solution

The solution is given by..


### Conclusion

The conclusion is...

# Section 2

Section 2 is about..

## Analyses

In this analyses..


```{r}
library(ggplot2)
bp <- ggplot(data=PlantGrowth, aes(x=group, y=weight, fill=group)) + geom_boxplot()
bp
```

```{r, warning = FALSE, error = FALSE, message=FALSE } 
library(plotly) 
library(ggplot2) 
library(dplyr) 

p <- ggplot(iris, aes(Sepal.Length, Petal.Length, colour=Species)) + geom_point() + theme_bw()
 ggplotly(p) 

library(leaflet) 

leaflet() %>% 
  addTiles %>% # Add default OpenStreetMap map tiles 
  setView(lng = 5.0, lat = 51.0, zoom = 6) 

``` 
