---
title: "Project 2: Master's thesis"
---

The project below describes parts of my Master's thesis to obtain my degree in Actuarial Science and Mathematical Finance. The thesis was nominated for the Johan de Witt award for the best actuarial thesis in 2016. 

The data set at hand consists of 163,660 records concerning a Belgian motor thirdparty 
liability (MTPL hereafter) insurance portfolio. All the observations 
relate to the year 1997. 

# Data

Reading in the data:

```{r, warning = FALSE, error=FALSE, message=FALSE}
library(readr)
MTPL_Data_1997 <- read_delim("MTPL_Data_1997.txt", "\t",
                             escape_double = FALSE, 
                             trim_ws = TRUE)
```

```{r, echo = FALSE, warning = FALSE, message=FALSE}
library(dplyr)
MTPL_Data_1997 <- MTPL_Data_1997 %>%
  select(NCLAIMS, AGEPH, AGEC, POWER, SPORT, CODPOSS)
```


The table below shows the covariates available in the MTPL insurance portfolio. 

```{r, warning = FALSE, fig.margin = TRUE}
knitr::kable(head(MTPL_Data_1997, 6))
```

Histograms for the variables given above:

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(dplyr)
belgieDat <- readRDS("belgieDat.rds")

inwoners <- readRDS("inwoners.rds") #read: niscode, gemeente, totaalbevolking, dichtheidperkm2, oppervlaktekm2
inkomen <- readRDS("inkomen.rds") #read: niscode, mediaaninkomen1000
data1997 <- readRDS("data1997.rds") #read: codposs, town, nclaims, amount, period, fuel
zipcodes <- readRDS("zipcodes.rds") %>% #read: niscode, codposs
  arrange(NISCODE) 
verkeer <- readRDS("verkeer.rds") #read: niscode, lengtekm, miljoenvoertuigkm

finaldata00 <- readRDS("finaldata.rds") #combination of data frames: inwoners, inkomen, data1997, zipcodes and verkeer
data1997a <- finaldata00 %>% 
  distinct(CODPOSS) %>% 
  left_join(., finaldata00) %>% 
  dplyr::select(CODPOSS, MEDIAANINKOMEN1000, ROAD.DENSITY, TRAFFIC.DENSITY.ST)

finaldata <- finaldata00 %>% 
  left_join(., belgieDat) %>% 
  arrange(ID_4) 

library(ggplot2)
library(ggthemes)
transparent_theme <- theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  axis.text.x = element_blank(), 
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  panel.grid = element_blank(),
  axis.line = element_blank(),
  panel.background = element_rect(fill = "transparent",colour = NA),
  plot.background = element_rect(fill = "transparent",colour = NA))

histboxplot <- function(mappinghist, xas, mapmean, data, mappingbox, xmin1, xmax1, width1, variable1, bins1){
  p1 <- ggplot(data = data, mappinghist) + 
        geom_histogram(aes(y = ..density..), bins = bins1, fill = "grey") + 
        geom_line(aes(y=..density..), stat = 'density', colour = "grey50", adjust = 2) +
        theme_bw() + 
        xlab(xas) + 
        ylab("Density") + 
        coord_cartesian() +
        geom_vline(mapmean, color = "grey50", linetype = 2) +
        scale_y_continuous(labels = function(x) format(x, nsmall = 3))
  p2 <- ggplot(data, mappingbox) + geom_tufteboxplot(colour = "steelblue", whisker.type = "line", median.type = "line", hoffset = 0, width = 5) + 
    coord_flip() + transparent_theme
  p2_grob <- ggplotGrob(p2)
  p1 + annotation_custom(grob = p2_grob, xmin = xmin1, xmax = xmax1, ymin = -width1, ymax = width1) + 
    scale_x_continuous(breaks = extended_range_breaks()(variable1))
}

histbox.claims <- histboxplot(aes(x = NCLAIMS), "Number of claims", aes(xintercept = mean(NCLAIMS)), finaldata, 
                              aes(y = NCLAIMS, x = factor(1)), min(finaldata$NCLAIMS), max(finaldata$NCLAIMS), 0.1, finaldata$NCLAIMS, 80) 

```

```{r, fig.show = 'hold', fig.height = rep(2,7), fig.cap="Histograms and Tufte boxplots for the variables in the MTPL data set", echo = FALSE}
histbox.claims <- histboxplot(aes(x = NCLAIMS), "Number of claims", aes(xintercept = mean(NCLAIMS)), finaldata, 
                              aes(y = NCLAIMS, x = factor(1)), min(finaldata$NCLAIMS), max(finaldata$NCLAIMS), 0.1, finaldata$NCLAIMS, 80) 
histbox.power <- histboxplot(aes(x = POWER), "Power vehicle (kW)", aes(xintercept = mean(POWER)), data1997, 
                             aes(y = POWER, x = factor(1)), min(data1997$POWER), max(data1997$POWER), 0.1, data1997$POWER, 70) 
histbox.ageph <- histboxplot(aes(x = AGEPH), "Age policyholder (years)", aes(xintercept = mean(AGEPH)), data1997, 
                             aes(y = AGEPH, x = factor(1)), min(data1997$AGEPH), max(data1997$AGEPH), 0.1, data1997$AGEPH, 79) 
histbox.agec <- histboxplot(aes(x = AGEC), "Age vehicle (years)", aes(xintercept = mean(AGEC)), data1997, 
                            aes(y = AGEC, x = factor(1)), min(data1997$AGEC), max(data1997$AGEC), 0.1, data1997$AGEC, 50) 
histbox.income <- histboxplot(aes(x = MEDIAANINKOMEN1000), "Median income (1000s euros)", aes(xintercept = mean(MEDIAANINKOMEN1000)), finaldata, 
                              aes(y = MEDIAANINKOMEN1000, x = factor(1)), min(finaldata$MEDIAANINKOMEN1000), max(finaldata$MEDIAANINKOMEN1000), 0.1, round(finaldata$MEDIAANINKOMEN1000,0), 50) 
histbox.density <- histboxplot(aes(x = ROAD.DENSITY), "Road density (road length in kilometers/area size in square kilometers)", aes(xintercept = mean(ROAD.DENSITY)), finaldata, 
                               aes(y = ROAD.DENSITY, x = factor(1)), min(finaldata$ROAD.DENSITY), max(finaldata$ROAD.DENSITY), 0.1, round(finaldata$ROAD.DENSITY, 0), 50) 
histbox.vehicles <- histboxplot(aes(x = MILJVOERTUIGKM), "Traffic density (number of vehicles/road length in kilometers)", aes(xintercept = mean(MILJVOERTUIGKM)), finaldata, 
                                aes(y = MILJVOERTUIGKM, x = factor(1)), min(finaldata$MILJVOERTUIGKM), max(finaldata$MILJVOERTUIGKM), 0.1, round(finaldata$MILJVOERTUIGKM, 0), 50)

histbox.claims
histbox.power
histbox.ageph
histbox.agec
histbox.income
histbox.density
histbox.vehicles

```

